Projit: An Open Source tool for Decoupled Data Science

### 2023-07-27 docs/paper/paper.tex
Resolving conflicts in paper body

\begin{abstract} 
Scientific practice has expanded to become increasingly dependent of digitial technologies,
large scale data processing and advanced analytical methods. These shifts have demanded new
methods of implementing and recording the details of scientific projects. Monolothic applications
have the advantage of a single and consistent design, however they impede the ability of users to
innovate and incrementally improve processes. We discuss the qualities of an ideal eScience framework
for building multi-stage collaboration scientific workflows and present an open source implementation
for managing data science processes in a decoupled fashion that permits both flexible implementation of
any stage of processing, and greater ease of meta-data analysis.
\end{abstract}

@@

Many Machine Learning experimentation frameworks focus on the task of making experiments
easier to execute and deploy into production systems\cite{Alberti:2018,MolnerDomenech:2020}. 
To do so they are often constructed for specific combinations of technology. This is a necessary limitation 
to enable efficiency, but has the effect of limiting general applicability.
Many other eScience frameworks focus on the lineage and management of data, referred to as 
the so-called provenance problem \cite{Sahoo:2008,Conquest:2021}
The goal of the provenance frameworks is sufficient auditibiliy of data that will 
render eScience transparent and repeatable. This can be auditing of data from multiple source systems,
auditing of logs generated during data processing\cite{Ferdous2020}. At the extreme we can seek to
quantify every transformation that happens to data in the course of processing\cite{Sahoo2009}. 
Regardless of the
specific data to be audited, these frameworks focus on developing unified systems and processes so
that auditing can be easily performed over many projects.

In addition to systems for storage of data, eScience application may include facilities for orchestration
of data processes and external services\cite{Subramanian2013}, requests for experiments with specific 
parameters\cite{Hunter:2005}, or integrated analysis of results, generation of insights and documentation. 
Other frameworks and approaches in eScience focus on understanding how to do large scale collaborative science, or
facilitate meta-level learning of various kinds\cite{Hunter:2005,Liu:2023}. The better we track the
process of science as a whole, the better we can understand both how to improve scientific processes
as well as data mine the history of experimental results for phenomena that were difficult to detect.

Frameworks for eScience will typically need to take a position on the extent to which they are domain
specific, versus general purpose. A domain specific approach that integrates multiple data sources in
a domain aware fashion can faciliate automated or assisted scientific discover\cite{Howe2008}. On the
other hand a general purpose framework facilitates multi-disciplinary collaboration and permits meta-analysis
that transcends the boundaries of disciplines. The other key dimension for a decision is the extent to
which an eScience application depends on specific technologies, many machine learning science platforms
can provide efficiency gains, but only when using specfic libraries and frameworks\cite{Alberti:2018,MolnerDomenech:2020}.
Similarly, other empirical science platforms are built on specific database, webserver or application
frameworks, which make them less extensible and harder to integrate.

In this work we argue for development of data science frameworks that are minimal in expectations, both in
terms of appplication domains and underlying technologies. We present a design framework for building
decoupled data science tools that can improve efficiency and replication through standardisation, without
unreasonable impositions on design decisions. We describe the design of a open source project integration
tool (\textit{projit}) that can be used either as a CLI or python API. Internally \textit{projit} depends 
only on a metadata store that uses the general purpose JSON format. 
As such it is trivial for developers to build interfaces in other
languages, or devise web service APIs for decentralised versions. We explore a case study of comparing results
across multiple projects for which we have used the \textit{projit} application to manage our metadata.

@@

\begin{itemize}
 \item Sources: Provenance of Data Sources
 \item Processing: Record of Data Processing
 \item Reuse: Facilitating Reuse of Datasets
 \item Tracking: Tracking of Experiments and outputs
 \item Results: Comparison of Methods and Results
 \item Documentation: Generation of Documentation
 \item Reproducibility: Facilitate reproduction of results
 \item Meta-Analysis: Facilitation of Meta-Analysis
\end{itemize}

@@

First and foremost, data driven projects require a access to the required \textbf{source} data 
and need to maintain records of this data provenance for \textbf{reproducibility}. 
There will typically be \textbf{processing} applied to these datasets to 
render them applicable to experimentation and analysis. An ideal tool will track the sequential 
nature of this \textbf{processing} as well as store information about the location of each resulting dataset.
The data processed in this way is then available for \textbf{reuse} across experiments and analysis,
making \textbf{results} comparable.

@@

In Figure \ref{fig:projit} we see that the core steps of data preparation, experimentation
and analysis of results all happen independently. Each of them accesses the projit metadata 
store for the information they need, and subsequently store information and results once complete.
This process means that in principle the location of an underlying dataset could change without
modifying other elements of the project. Similarly, we might change the parameters or an experiment
or the set of metrics we calculate. Each experiment and analysis task operates independently of the
others, and all that suffers when changes are made is the potential for comparison across equal
dimensions of variation.

In addition to the dominant requirements of experimentation (parameter and results) we store the results
of each experimental execution as well as the experiment duration, measured from the time of initiation
to completion. These records are particularly important in data science and machine learning where we
may want to trade off performance with computational requirements. But these values could be used to store
information about real world experimtal execution, or the time required to marshal and sequence a series
of independent web services.

@@

Projit has been implemented as python package that functions as both a command line application
and library that can be included inside other scripts and applications. The command line application
can be used to query the project metadata in much the same way that the git application can be used.
A user can add, modify and list the collection of analytical assets in the project: datasets, experiments 
parameters, tags and results are all accessible from the command line application.

The python package can be included in a script so that the script can access the project metadata store.
This allows the script to find the location of common datasets, register themselves as an experiment
and store results once the script is complete. Programmatic interaction with the project data through
the projit API is what permits the scripts of a project to be decoupled and contribute to the project
without being aware of how any other element is structured or implemented. Furthermore, as the metadata
is stored locally in standard JSON files, these can be stored inside a cloud repository and then contributed
to by collaborators. This allows distributed data science teams to define their own lines of experimentation,
but use a synchronised data set then continually contribute to a central meta-data store of project results.

@@

We have utilised the projit application across multiple data science projects to store reusable datasets
and the results of all experiments. Additionally, the metadata store contains infomation about the number
of times each experiment has been executed, and the execution time utilised on each run. This allows us
to generate an ad hoc script that can compare projects in terms of the data used, the number of experiments
conducted and the total execution time. This script is constructed for illustrative purposes to show that
the projit tool can permit arbitrary meta-analysis of projects through the standardised metadata stored across
git repositories.

>>>>

\begin{abstract} 
Scientific practice has expanded to become increasingly dependent of digitial technologies,
large scale data processing and analytical methods for working with large data volumes. 
These shifts have demanded new methods of implementing and recording the details of 
scientific projects. Monolothic applications have the advantage of a single and 
consistent design, however they impede the ability of users to innovate and 
incrementally improve processes. We discuss the qualities of an ideal eScience 
framework for building multi-stage collaboration scientific workflows and present 
an open source implementation for managing scientific processes in a decoupled 
fashion that permits both flexible implementation of any stage of processing, 
and greater ease of meta-data analysis.
\end{abstract}

@@

Many Machine Learning experimentation frameworks focus on the task of making experiments
easier to execute and deploy into production systems\cite{Alberti:2018,MolnerDomenech:2020}. 
To do so they are often constructed for specific combinations of technology. 
This is often necessary limitation to either make a project feasible or enable efficiency, 
but it has the side-effect of limiting general applicability.
Many other eScience frameworks focus on the lineage and management of data, referred to as 
the so-called provenance problem \cite{Sahoo:2008,Conquest:2021}.
The goal of the provenance frameworks is sufficient auditibiliy of data that will 
render eScience transparent and repeatable. This can be auditing of data from multiple 
source systems, or auditing of logs generated during data processing\cite{Ferdous2020}. 
At the extreme we can seek to quantify every transformation that happens to data in 
the course of processing\cite{Sahoo2009}. Regardless of the
specific data to be audited, these frameworks focus on developing unified systems and 
processes so that auditing can be easily performed over many projects.

In addition to systems for storage of data, eScience application may include facilities
for orchestration of data processes and external services\cite{Subramanian2013}, 
requests for experiments with specific parameters\cite{Hunter:2005}, 
or integrated analysis of results, generation of insights and documentation. 
Other frameworks and approaches in eScience focus on understanding how to do large 
scale collaborative science, or facilitate meta-level learning of various 
kinds\cite{Hunter:2005,Liu:2023}. The better we track the
process of science as a whole, the better we can understand both how to improve 
scientific processes as well as data mine the history of experimental results for 
phenomena that are difficult to detect.

Frameworks for eScience will typically need to take a position on the extent to which 
they are domain specific versus general purpose. A domain specific approach can 
integrate multiple data sources in a domain aware fashion that can faciliate 
automated or assisted scientific discover\cite{Howe2008}. On the
other hand a general purpose framework facilitates multi-disciplinary collaboration 
and permits meta-analysis that transcends the boundaries of disciplines. 
The other key dimension for a decision is the extent to which an eScience 
application depends on specific technologies, many machine learning science platforms
can provide efficiency gains, but only when using specfic libraries and 
frameworks\cite{Alberti:2018,MolnerDomenech:2020}. Similarly, other empirical 
science platforms are built on specific database, webservers or application
frameworks, which make them less extensible and harder to integrate.

In this work we argue for development of data science frameworks that impose minimal 
requirements, both in terms of appplication domains and underlying technologies. 
We present a design framework for building decoupled data science tools that can 
improve efficiency and replication through standardisation, without
unreasonable impositions on design decisions. We describe the design of a open 
source project integration tool (\textit{projit}) that can be used either as a 
Command Line Application (CLI) or python API. Internally \textit{projit} depends 
only on a metadata store that uses the general purpose JSON format. 
As such it is trivial for developers to build interfaces in other
languages, or devise web service APIs for decentralised versions. We explore a 
case study of comparing results across multiple projects for which we have used 
the \textit{projit} application to manage our metadata.

@@

\begin{itemize}
 \item Sources: Provenance of Data Sources
 \item Processing: Records of Data Processing / Transformations
 \item Reuse: Tooling to Facilitate Reuse of Datasets
 \item Tracking: Tracking of Experiments and Outputs
 \item Results: Comparison of Methods and Results
 \item Documentation: Generation of Project Documentation
 \item Reproducibility: Facilitation of reproduction of results
 \item Meta-Analysis: Facilitation of Multi-Project Meta-Analysis
\end{itemize}

@@

First and foremost, data driven projects require access to the required \textbf{source} data 
and need to maintain records of the data provenance for \textbf{reproducibility}. 
There will typically be \textbf{processing} applied to these datasets to 
render them applicable to experimentation and analysis. An ideal tool will track the 
sequential nature of this \textbf{processing} as well as store information about 
the location of each resulting dataset. The data processed in this way is then 
available for \textbf{reuse} across experiments and analysis,
making \textbf{results} comparable.

@@

In Figure \ref{fig:projit} we see that the core steps of data preparation, experimentation
and analysis of results all happen independently. Each of them accesses the projit metadata 
store for the information they need, and subsequently store information and results once 
complete. This process means that in principle the location of an underlying dataset 
could change without modifying other elements of the project. Similarly, we might change 
the parameters or an experiment or the set of metrics we calculate. Each experiment and 
analysis task operates independently of the others, relying on the meta-data store for 
dependencies rather than direct communication. When changes are made there might be a
loss of synchronisation between elements, for example between test data and results. 
This issue can only be addressed by extending the project with a dependency graph for 
capturing, and potentially orchestrating, processing pipelines.

In addition to the dominant requirements of experimentation (parameters and results) we 
store the results of each experimental execution as well as the experiment duration, 
measured from the time of initiation to completion. These records are particularly 
important in data science and machine learning where we may want to trade off 
performance with computational requirements. But these values could be used to store
information about real world experimental execution or the time required to marshall 
and sequence a series of independent web services.

The project utilises a notion of key asset types: dataset, experiment and results. Each
of these can be created independently with a reference between them in a loose coupling.
In addition, a tagging system allows these assets to be labelled with an open ended number
of tags that cen be used when extracting or tabulating project properties.

@@

Projit has been implemented as python package that functions as both a command line 
application and library that can be included inside other scripts and applications. 
The command line application can be used to query the project metadata in much the 
same way that the git application can be used to manipulate a query a set of software
source files. A user can add, modify and list the collection of assets in the 
project: datasets, experiments and results are all accessible from the command line 
application.

The python package can be included in a script so that the script can access the 
project metadata store. This allows the script to find the location of common 
datasets, register themselves as an experiment or execution and store results once 
the script is complete. Programmatic interaction with the project data through
the projit API is what permits the scripts of a project to be decoupled and 
contribute to the project without being aware of how any other element is 
structured or implemented. Furthermore, as the metadata
is stored locally in standard JSON files, these can be stored inside a cloud 
repository and then contributed to by collaborators. 
This allows distributed data science teams to define their own lines of experimentation,
but use a synchronised data set then continually contribute to a central meta-data 
store of project results.

@@

We have utilised the projit application across multiple data science projects to store 
reusable datasets and the results of all experiments. Additionally, the metadata store 
contains information about the number of times each experiment has been executed, 
and the execution time utilised on each run. This allows us to generate an ad hoc 
script that can compare projects in terms of the data used, the number of experiments
conducted and the total execution time. This script is constructed for illustrative 
purposes to show that the projit tool can permit arbitrary meta-analysis of projects 
through the standardised metadata stored across git repositories.
